A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-04-15 03-semantics.tex
Minor rewrites for brevity and clarity in the beginning of Section 3

In this section, we introduce a reduced syntax for \CEU and propose an
operational semantics to formally describe the behavior of its programs.
We describe a small synchronous kernel highlighting the peculiarities of \CEU,
in particular, the stack-based execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language,
leaving out side-effects and system calls (which behave like in conventional
imperative languages).

@@

The grammar below defines the syntax of a subset of \CEU that is
sufficient to describe all semantic peculiarities of the language.
\bgroup
\def\lbl#1{\quad\text{\emph{#1}}}%
\newdimen\X
\X=-1\jot
\begin{alignat*}{2}
  P\Coloneqq
      &\enspace\ceu{\Mem(\Id)}
      &&\lbl{any memory access to~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\AwaitExt(\Id)}
      &&\lbl{await external event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\AwaitInt(\Id)}
      &&\lbl{await internal event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\EmitInt(\Id)}
      &&\lbl{emit internal event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\Break}
      &&\lbl{loop escape}\\[\X]
      %%
  \mid&\enspace\ceu{\IfElse{\Mem(\Id)}{P_1}{P_2}}
      &&\lbl{conditional}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\,;\,P_2}
      &&\lbl{sequence}\\[\X]
      %%
  \mid&\enspace\ceu{\Loop P_1}
      &&\lbl{repetition}\\[\X]
      %%
  \mid&\enspace\ceu{\Every{\Id}\ P_1}
      &&\lbl{event iteration}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\And P_2}
      &&\lbl{par/and}\\[\X]
      %%
  \mid&\enspace\ceu{P_2\Or P_2}
      &&\lbl{par/or}\\[\X]
      %%
  \mid&\enspace\ceu{\Fin P}
      &&\lbl{finalization}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\AtLoop P_2}
      &&\lbl{unwinded loop}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\AtAnd\ P_2}
      &&\lbl{unwinded par/and}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\AtOr\ P_2}
      &&\lbl{unwinded par/or}\\[\X]
      %%
  \mid&\enspace\ceu{\CanRun(n)}
      &&\lbl{can run on stack level~$n$}\\[\X]
      %%
  \mid&\enspace\ceu{\Nop}
      &&\lbl{terminated program}
\end{alignat*}
\egroup

The~$\ceu{\Mem(id)}$ primitive represents all accesses, assignments, and system
calls that affect a memory location identified by~$id$.
%
According to the synchronous hypothesis of \CEU, $\ceu{\Mem}$ expressions are
considered to be atomic and instantaneous.
%
As the challenging parts of \CEU reside on its control structures, we are not
concerned here with a precise semantics for side effects, but only with their
occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an
%unused identifier).

We assume that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$
and~$\ceu{\EmitInt}$ expressions do not share identifiers:
any identifier is either a variable, an external event, or an internal event.

Most expressions in the abstract language are mapped to their counterparts in
the concrete language.
The exceptions are the finalization block~$\ceu{\Fin{P}}$ and the
\texttt{@}-expressions which result from expansions in the transition rules to
be presented.

Regarding mismatches between the concrete and abstract languages, the concrete
\code{await} and \code{emit} primitives support communication of values between
them, e.g., an ``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
variable~\code{v} to~10.
To reproduce this functionality in the formal semantics, we can use a shared
variable to hold the value of an $\ceu{\EmitInt}$ and access it after the
corresponding $\ceu{\AwaitInt}$ awakes.
%
Also, a ``\code{finalize $A$ with $B$ end; $C$}'' in the concrete language is
equivalent to ``\ceu{A;\;((\Fin{B})\ \Or\ C)}'' in the abstract language.
In the concrete language, $A$ and~$C$ execute in sequence, and
the finalization code~$B$ is implicitly suspended waiting for~$C$
termination.
In the abstract language, ``$\ceu{\Fin B}$'' suspends forever when reached (it is
an awaiting expression that never awakes).
Hence, we need an explicit \code{or} to execute~$C$ in parallel, whose
termination aborts ``$\ceu{\Fin B}$'', which finally causes~$B$ to
execute (by the semantic rules to be discussed).

@@

The core of our semantics describes how a program reacts to a single external
input event, i.e., starting from an input event, how the program behaves and
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules~\cite{Plotkin-G-D-1981} which
are designed in such a way that at most one transition is possible at any
time, resulting in deterministic reactions.
%
The transition rules map a triple with a program~$p$, a stack level~$n$, and an
emitted event~$e$ to a modified triple as follows:
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are nonnegative
integers representing the current stack level, and~$e,e'\in\E\cup\{\nil\}$ are
the events emitted before and after the transition (both possibly being the
empty event~$\nil$).
$\E$ is a set of primitive events.

We will refer to the triples on the left-hand and right-hand sides of
symbol~$\trans$ as \emph{descriptions} (denoted~$\delta$).  The triple on the
left-hand side of symbol~$\trans$ is called the \emph{input description}, and
the triple on its right-hand side is called the \emph{output description}.

@@

At the beginning of a reaction to an input event~$id$, the input description is
initialized with stack level~0 ($n=0$) and with the externally emitted event
($e=id$).
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the last output description will block with a (possibly) modified program~$p'$, at stack
level~0, and with no event emitted~($\nil$):
\[
  \<p,0,e>\mathbin{\trans[*]}\<p',0,\nil>\,.
\]

>>>>

In this section, we introduce a reduced, abstract syntax for \CEU and
present an operational semantics that formalizes the behavior of its
programs.
% We describe a small synchronous kernel highlighting the peculiarities of \CEU,
% in particular, the stack-based execution for internal events.
%
The semantics deals only with the control aspects of \CEU.
Side-effects and system calls are encapsulated in an abstract memory access
primitive and are assumed to behave like in conventional imperative
languages.

@@

The grammar below defines the syntax of a subset of \CEU that is
sufficient to describe all peculiarities of the language.
\bgroup
\def\lbl#1{\quad\text{\emph{#1}}}%
\newdimen\X
\X=-1.25\jot
\begin{alignat*}{2}
  P\Coloneqq
      &\enspace\ceu{\Mem(\Id)}
      &&\lbl{any memory access to~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\AwaitExt(\Id)}
      &&\lbl{await external event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\AwaitInt(\Id)}
      &&\lbl{await internal event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\EmitInt(\Id)}
      &&\lbl{emit internal event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\Break}
      &&\lbl{loop escape}\\[\X]
      %%
  \mid&\enspace\ceu{\IfElse{\Mem(\Id)}{P_1}{P_2}}
      &&\lbl{conditional}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\,;\,P_2}
      &&\lbl{sequence}\\[\X]
      %%
  \mid&\enspace\ceu{\Loop P_1}
      &&\lbl{repetition}\\[\X]
      %%
  \mid&\enspace\ceu{\Every{\Id}\ P_1}
      &&\lbl{event iteration}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\And P_2}
      &&\lbl{par/and}\\[\X]
      %%
  \mid&\enspace\ceu{P_2\Or P_2}
      &&\lbl{par/or}\\[\X]
      %%
  \mid&\enspace\ceu{\Fin P}
      &&\lbl{finalization}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\AtLoop P_2}
      &&\lbl{unwinded loop}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\AtAnd\ P_2}
      &&\lbl{unwinded par/and}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\AtOr\ P_2}
      &&\lbl{unwinded par/or}\\[\X]
      %%
  \mid&\enspace\ceu{\CanRun(n)}
      &&\lbl{can run on stack level~$n$}\\[\X]
      %%
  \mid&\enspace\ceu{\Nop}
      &&\lbl{terminated program}
\end{alignat*}
\egroup

The~$\ceu{\Mem(id)}$ primitive represents any access, assignment, or system
call that affects the memory location identified by~$id$.\footnotemark\
%
Following the synchronous hypothesis, $\ceu{\Mem}$ expressions are
considered to be atomic and instantaneous.
%
% As the challenging parts of \CEU reside on its control structures, we are not
% concerned here with a precise semantics for side effects, but only with their
% occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an
%unused identifier).
%
\footnotetext{Although the same symbol~$\ceu{\Id}$ is used in their
  definition, the $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$
  and~$\ceu{\EmitInt}$ expressions do not share identifiers: any identifier
  is either a variable, an external event, or an internal event.}

Most expressions in the abstract language are mapped to their counterparts
in the concrete language.  The exceptions are the finalization
block~$\ceu{\Fin{p}}$ and the \texttt{@}-expressions which do not exist in
the concrete language.  (These expressions result from the expansion of the
transition rules to be presented.)

A further difference between the concrete and abstract languages regards the
emit-await pair.  In the concrete language the \code{await} and \code{emit}
expressions support communication of values between them, e.g., an
``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
variable~\code{v} to~10.  Although the abstract $\ceu{\AwaitInt}$
and~$\ceu{\EmitInt}$ expression do not support such communication of values,
it can be easily simulated: we can use a shared variable to hold the value
of an $\ceu{\EmitInt}$ and access it after the corresponding
$\ceu{\AwaitInt}$ awakes.

Finally, a ``\code{finalize $A$ with $B$ end; $C$}'' in the concrete
language is equivalent to ``\ceu{A;\;((\Fin{B})\ \Or\ C)}'' in the abstract
language.  In the concrete language, $A$ and~$C$ execute in sequence, and
the finalization code~$B$ is implicitly suspended waiting for~$C$
termination.  In the abstract language, ``$\ceu{\Fin B}$'' suspends forever
when reached (it is an awaiting expression that never awakes).  Hence, we
need an explicit \code{or} to execute~$C$ in parallel, whose termination
aborts ``$\ceu{\Fin B}$'', which finally causes~$B$ to execute (by the
semantic rules below).

@@

The core of our semantics describes how a program reacts to a single external
input event, i.e., starting from an input event, how the program behaves and
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules~\cite{Plotkin-G-D-1981} which
ensure that at most one transition is possible at any time, resulting in
deterministic reactions.
%
The transition rules map a triple with a program~$p$, a stack level~$n$, and an
emitted event~$e$ to a modified triple as follows:
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are nonnegative
integers representing the current stack level, and~$e,e'\in\E\cup\{\nil\}$ are
the events emitted before and after the transition (both possibly being the
empty event~$\nil$).
%$\E$ is a set of primitive events.

We will refer to the triples on the left-hand and right-hand sides of
symbol~$\trans$ as \emph{descriptions} (denoted~$\delta$).  The description
on the left-hand side of~$\trans$ is the \emph{input description},
and the one on the right-hand side is the \emph{output description}.

@@

At the beginning of a reaction to an input event~$id$, the input description is
initialized with stack level~0 ($n=0$) and with the externally emitted event~$e=id$.
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the last output description will block with a (possibly) modified program~$p'$ at stack
level~0 and with no event emitted:
\[
  \<p,0,e>\mathbin{\trans[*]}\<p',0,\nil>\,.
\]
